Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-17 manuscript/paper.tex
intro completed

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
octahedra
 in ABO$_3$ perovskites
and study the
effect of the crystal field of the nearest A and B ion cages 
%that immediately surround the octahedra
on the single electron spectrum of the 
central 
transition metal ions.
While the crystal field of the octahedra 
that 
% quenches the orbital angular momentum of the central B ion and 
creates $e_g$ and $t_{2g}$ manifolds
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field,
leading to sizeable splittings in the $e_g$ and $t_{2g}$ manifolds of the central B ion.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the local orbital ordering are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

\com{
para1: 
perovskites .. imp\\
octa rot imp...

para2: rot understood in terms of a tolerance factor... relative sizes of ions...\\
nothing substantial exists in the literature to explore the cause/reasons ... rotations...\\
except only recently, nick paper,... hydrogen bonding etc...

para3:
JT, ...
}

Perovskites are
technologically important materials 
that exhibit a rich variety of electronic, magnetic and optical properties
due to the interplay between \az{[refine...]}
orbital, spin, correlations, charge and structure.

\az{
orbital ordering... 
orbital-structure interplay.... structure-property relationships... 
}

@@

Their crystal structures have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can distort or simply rotate to lower the symmetry
and produce a tetragonal or orthorhombic structure.
While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller effect
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

@@

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL2010,CammarataJCP2014,nick2016,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez et. al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL2010}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP2014}.
% Lee, nick2016
Lee et. al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{nick2016}.
Yoshida et. al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

We take a completely different approach to this problem.
We go back to the basics
and see 
what is it that lifts 
 the orbital degeneracy in the classical JT effect?
It is a specific 
deformation of the crystal field 
%(${\sim z^2}$)
at the transition metal (JT ion) site.
Does the same happen in the case of octahedral tilting and rotations?
While the potential of an octahedral anion cage 
in its own frame
does not change 
%in case of 
with
rigid rotations for obvious reason,
the total crystal field 
does deform indeed
because
 the rest of the lattice is not only rotated
  in a given octahedron's frame (that turns its non-spherical potential)
 but also distorted owing to the accompanying structural transformation.
%So, in principle, there should be a (first order) JT effect at play.
Using this insight
and elements of
crystal field theory,
here in this manuscript
we explore this problem in detail
and find that,
just like a first order JT effect,
 the deformed crystal field 
in this case also lifts the orbital degeneracy,
by a significant amount for typical system parameters,
signalling the instability against the rotations.

@@

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B.
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
the splittings in/of? its $t_{2g}$ and $e_g$ manifolds,
created primarily by the anion's octahedral potential.
We find that the size of the splitting 
and local orbital ordering
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

\az{
structure of the paper:\\
}

@@

\com{
What is it that lifts the degeneracy in the classical JT effect?
its a deformation of the crystal field at the transition metal/JT ion site.
Does a deformation happen in the case of octahedral tilting and rotations?
Ridig octahedral.... octaherdal field of the octahedral anions cage is not changed 
... but... AB cages and the rest of the lattice is rotated and deformed... so... of course the CF should deform...
So there are in principle clear signs that there should be a (first order) JT effect at play.
}

\az{
Here, we .... just like the classical JT effect, where octahedral potential deforms....
....

we consider Pbnm.... other tilting and rotation.. glazer notation...???
}

\rev{structure of the paper:\\

\section{Formalism: Crystal field Hamiltonian using multipole expansion of the crystal field potential}

@@

\section{Janh-Teller Effect}

@@

\section{Model and calculations of ABO$_3$ Perovskites}%}

@@

\subsection{Rigid octahedral rotations}

\subsection{Transformation of the crystal structure}

@@

\subsection{Deformation of the crystal field}

@@

\section{Results: Single particle spectrum and JT instability}

@@

\subsection{$e_{g}$ manifold}

@@

\subsection{Effect of the full lattice}

@@

\section{Appendix: Rotation matrices}

>>>>

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
octahedra
 in ABO$_3$ perovskites
and study the
effect of the crystal field of the nearest A and B ion cages 
%that immediately surround the octahedra
on the single electron spectrum of the 
central 
transition metal ions.
While the crystal field of the octahedra 
that 
% quenches the orbital angular momentum of the central B ion and 
creates $e_g$ and $t_{2g}$ manifolds
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in the $e_g$ and $t_{2g}$ manifolds of the central B ion.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the local orbital ordering are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

Perovskites  
exhibit a rich variety of intriguing 
electronic, magnetic and optical properties
due to the interplay between
spin, orbital, and structural degrees of freedom
and widely varying electronic correlations.
These materials 
are important
both for practical applications in all sorts of modern devices 
and 
fundamental understanding of various physical phenomena.

\com{
Perovskite?s potential applications are varied and include uses in sensors and catalyst electrodes, certain types of fuel cells, solar cells, lasers, memory devices and spintronics applications.
}

@@

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can distort or simply rotate to lower the symmetry
and produce a tetragonal or orthorhombic structure.
While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller effect
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

@@

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL2010,CammarataJCP2014,nick2016,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL2010}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP2014}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{nick2016}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

We take a completely different approach to this problem.
We go back to the basics
and see 
what is it that lifts 
 the orbital degeneracy in the classical JT effect?
It is a specific 
deformation of the crystal field 
%(${\sim z^2}$)
at the transition metal (JT ion) site.
Does the same happen in the case of octahedral tilting and rotations?
While the potential of an octahedral anion cage 
in its own frame
does not change 
%in case of 
with
rigid rotations for obvious reason,
the total crystal field 
does deform indeed
because
 the rest of the lattice is not only rotated
  in a given octahedron's frame (that turns its non-spherical potential)
 but also distorted owing to the accompanying structural transformation.
%So, in principle, there should be a (first order) JT effect at play.
Using this insight
and elements of crystal field theory,
here in this manuscript,
we 
consider next two coordination cages around an octahedron
% explore this problem in detail
and find that,
just like a first order JT effect,
 the deformed crystal field 
in this case also lifts the orbital degeneracy,
by a significant amount for typical system parameters,
signalling the instability against the rotations.

@@

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B.
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
the splittings in its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the size of the splitting 
and local orbital ordering
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

@@

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages,
in the cubic structure, to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations (sec.~\ref{sec:rotation})
on
the structure (sec.~\ref{sec:structure}) 
and crystal field (sec.~\ref{sec:deformCF}).
The results are in sec.~\ref{sec:results},
where we seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (sec.~\ref{sec:t2g})
and $e_g$ (sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (sec.~\ref{sec:FL}).
\az{appendix?}

\section{Formalism: Crystal field Hamiltonian using multipole expansion of the crystal field potential}
\label{sec:formalistm}

@@

\section{Janh-Teller Effect}
\label{sec:JT}

@@

\section{Model and calculations of ABO$_3$ Perovskites}%}
\label{sec:calc}

@@

\subsection{Rigid octahedral rotations}
\label{sec:rotation}

\subsection{Transformation of the crystal structure}
\label{sec:structure}

@@

\subsection{Deformation of the crystal field}
\label{sec:deformCF}

@@

\section{Results: Single particle spectrum and JT instability}
\label{sec:results}

@@

\subsection{$e_{g}$ manifold}
\label{sec:eg-gen}

@@

\subsection{Effect of the full lattice}
\label{sec:FL}

@@

\section{Appendix: Rotation matrices}
\label{sec:append}
